---
title: "HW8 R"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Stat 322 HW8 R problems

1. We are interested in testing $H_0: \mu=60$ vs. $H_1: \mu \neq 60$. Initially, we consider a size .05 test on a random sample of size 16 from a normal distribution with $\sigma=4$ .  

```{r}
sigma <- 4
n <- 16
alpha <- .05
mu0 <- 60
mu <- seq(50,70,length=1000)
power <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
```

  a. generate a plot showing the effect of sample size on power (i.e. superimpose the graphs of three power functions representing 3 different sample sizes:$n=16, n>16,$ and $n<16$)
     
```{r}
plot(mu,power,type="l")
n <- 8
power2 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power2,lty=2)
n <- 32
power3 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power3,lty=4)
n <- 16
```

  b. generate a plot showing the effect of test size on power (i.e. superimpose the graphs of three power functions representing 3 different test sizes: $\alpha(\delta)=.05, \alpha(\delta) > .05, \alpha(\delta) < .05$)

```{r}
plot(mu,power,type="l")
alpha <- .10
power2 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power2,lty=2)
alpha <- .01
power3 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power3,lty=4)
alpha <- .05
```
     
  c. generate a plot showing the effect of standard deviation on power (i.e. superimpose the graphs of three power functions representing 3 different standard deviations: $\sigma=4, \sigma=4, \sigma=4$)
     
```{r}
plot(mu,power,type="l")
sigma <- 8
power2 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power2,lty=2)
sigma <- 2
power3 <- 1 - pnorm(qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma) +
  pnorm(-qnorm(1-alpha/2,0,1) + (mu0-mu)*sqrt(n)/sigma)
lines(mu,power3,lty=4)
sigma <- 4


```

2. Based on the description for Exercise 9.1.3:
  a. Sketch the power function described in (a).
```{r}
p <- seq(0,1,by=.01)
power <- 1 - pbinom(6,size=20,prob=p) + pbinom(1,size=20,prob=p)
plot(p,power,type="l")
```
     
  b. If we desire a level .05 test for $H_0: p=0.2$ vs $H_1: p>0.2$, define the appropriate critical region.  What is the size of this test?
```{r}
y <- 1:20
probtype1error <- 1 - pbinom(y-1,size=20,prob=.20)
cbind(y,probtype1error)
```
Reject Ho if Y>=8; size of test = .032

3. Duplicate the two plots in Figure 9.12 (p. 580)'
```{r}
psi <- seq(0,6,length=1000)
alpha <- .05
degfr <- c(1,2,3,5,10,60)
power <- 1 - pt(qt(1-alpha,df=degfr[1]),df=degfr[1],ncp=psi)
plot(psi,power,type="l",ylim=c(0,1))
for (i in 2:6)  {
  power <- 1 - pt(qt(1-alpha,df=degfr[i]),df=degfr[i],ncp=psi)
  lines(psi,power,lty=i)  }

alpha <- .01
degfr <- c(1,2,3,5,10,60)
power <- 1 - pt(qt(1-alpha,df=degfr[1]),df=degfr[1],ncp=psi)
plot(psi,power,type="l",ylim=c(0,1))
for (i in 2:6)  {
  power <- 1 - pt(qt(1-alpha,df=degfr[i]),df=degfr[i],ncp=psi)
  lines(psi,power,lty=i)  }
```

