---
title: "Image Compression"
author: "Matt Richey"
date: "4/27/2020"
output:
  pdf_document: default
  html_document: default
---


# Setup and Load Libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
```
Load libraries
```{r}
suppressMessages(library(tidyverse))
```

# Introduction
This a little off the beaten path, but it's fun to see how SVD  can be used for data compression.


# Data
Load an image file of greyscale  values.
```{r}
dataDir <- "/Users/richeym/Dropbox/COURSES/ADM/ADM_S20/CODE/Chap10"
dataFile <- "Mandela.csv"
mandela.mat <- read.csv(file.path(dataDir,dataFile))
mandela.mat <-  scale(mandela.mat,scale=F)
dim(mandela.mat)
```
Hence we have a 400 x 468 dimenional image of Nelson Mandela.

```{r}
image(mandela.mat,col=grey.colors(256))
```

Perform a SVD decomposition
```{r}
man.svd <- svd(mandela.mat)
```

Here are the u, d, and v matrices
```{r}
U <- man.svd$u
D <- man.svd$d
V <- man.svd$v
dim(U)
```



For any value k=1,2,...64 we can create the best "rank k" approximation to rectCat by only using k of the orthonormal bases U and V and the first k columns of the diagonal matrix D.


Let's try k=10. 
```{r}
k <- 10
compMan <-  U[,1:k] %*% diag(D[1:k]) %*% t(V[,1:k])
image(compMan,col=grey.colors(256))

```

Sorta blury, but note that

   * U[,1:k] and V[,1:k] are both is 400 x k
   * D[1:k,1:k] is has k nonzero entries
   
Hence the total amount of data needed for this is approximately 800k. For k=10, this is about 8000. The uncompressed image  requires 400*468 = [1] 187200 units.



Build a sequence of compressions. The last one is the full image
```{r}
kvals <- c(2,4,8,16,32,64,128,256,400)

op <- par(mfrow=c(3,3),mai=c(0.02,0.,0,0))
for(k in kvals){
 compMan <-  U[,1:k] %*% diag(D[1:k]) %*% t(V[,1:k])
image(compMan,col=grey.colors(256))
}
par(op)

```

By the time k=64, the image is fairly sharp. This image requires 800*64=51200 units. Compared with the original this represents only 27% of the size!


```{r}
k <- 64
compMan <-  U[,1:k] %*% diag(D[1:k]) %*% t(V[,1:k])
image(compMan,col=grey.colors(256))

```