---
title: "Bayes Test Procedure in R"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r,include=FALSE}
library(mosaic)
```
# TV Radiation Example

```{r}
# The data - TV radiation example
x <- c(0.65,0.28,0.47,0.44,0.25,1.03,0.28,0.64,0.71,0.49)
n <- length(x)
xbar <- mean(x)
snsq <- (n-1)*var(x)
```

$H_0:\mu \leq \mu_0$ vs $H_1:\mu > \mu_0$

Consider Pr($H_0$ true | data) and Reject $H_0$ if small. $Pr(H_0 \ true | x)=\int_{\Omega_0}\xi(\mu|x)d\mu$


# Non-informative Prior:

$\xi(\mu,\tau)=\xi(\mu|\tau)\xi(\tau)=N(\mu_0,\lambda_0\tau) gamma(\alpha_0,\beta_0)$

For what values of $\mu_0, \lambda_0, \alpha_0, \beta_0$ would $\xi(\mu,\tau)=1/\tau$ (Jeffreys Prior)?

```{r}
# Usual improper prior hyperparameters
mu0 <- 0
lambda0 <- 0
alpha0 <- -0.5
beta0 <- 0
```


2. Find the values of $\mu_1, \lambda_1, \alpha_1, \beta_1$ using the non informative prior.

```{r}
mu1 <- (lambda0*mu0+n*xbar)/(lambda0+n)
lambda1 <- lambda0 + n
alpha1 <- alpha0 + (n/2)
beta1 <- beta0 + (snsq/2) + ((n*lambda0*(xbar-mu0)^2)/(2*(lambda0+n)))

mu1;lambda1;alpha1;beta1
```

3. Using these posterior values show the equivalence of the Bayes Test Procedure using Jeffreys prior to the frequentist result ($U\sim t_{(n-1)}$)

```{r}
# Marginal posterior for mu
mu <- seq(.20,.80,length=1000)
dmu <- mu[2]-mu[1]
U1 <- sqrt((lambda1*alpha1)/beta1)*(mu-mu1)
y1 <- dt(U1,2*alpha1)

# Posterior probability that Ho true
Uat5 <- sqrt((lambda1*alpha1)/beta1)*(0.50-mu1)
ht5 <- dt(Uat5,2*alpha1)/(sum(y1)*dmu)
probHotrue <- pt(Uat5,2*alpha1)
probHotrue
```


4. Compare the posterior probability to p-value (using R).

```{r}
t.test(x,mu=.5,var.equal=TRUE,alternative="greater")
```


5. Repeat using informative hyper parameters for the conjugate prior relationship. ($\mu_0 = .25; \lambda_0 = 4; \alpha_0 = 2; \beta_0 = 0.5$)

```{r}
# Informative prior hyperparameters
mu0 <- .25
lambda0 <- 4
alpha0 <- 2
beta0 <- 0.5
```

```{r}
# Posterior hyperparameters
mu1 <- (lambda0*mu0+n*xbar)/(lambda0+n)
lambda1 <- lambda0 + n
alpha1 <- alpha0 + (n/2)
beta1 <- beta0 + (snsq/2) + ((n*lambda0*(xbar-mu0)^2)/(2*(lambda0+n)))

# Marginal posterior for mu
U1b <- sqrt((lambda1*alpha1)/beta1)*(mu-mu1)
y1b <- dt(U1b,2*alpha1)

# Posterior probability that Ho true
Uat5b <- sqrt((lambda1*alpha1)/beta1)*(0.50-mu1)
ht5b <- dt(Uat5b,2*alpha1)/(sum(y1b)*dmu)
probHotrue <- pt(Uat5b,2*alpha1)
probHotrue
```

```{r}
# Superimposed posteriors for mu
plot(mu,y1b/(sum(y1b)*dmu),type="l",xlab="mu",ylab="density",
  xlim=c(.2,.8),
  ylim=c(0,max(y1b/(sum(y1b)*dmu),y1/(sum(y1)*dmu))),
  main="Posterior dstns of mu (Noninformative prior dashed)")
lines(mu,y1/(sum(y1)*dmu),lty=2)
lines(c(.5,.5),c(0,max(ht5,ht5b)))
```

6. Show equivalence with non informative prior to the two-sided t-test. 


```{r}
# Usual improper prior hyperparameters
mu0 <- 0
lambda0 <- 0
alpha0 <- -0.5
beta0 <- 0

# Posterior hyperparameters
mu1 <- (lambda0*mu0+n*xbar)/(lambda0+n)
lambda1 <- lambda0 + n
alpha1 <- alpha0 + (n/2)
beta1 <- beta0 + (snsq/2) + ((n*lambda0*(xbar-mu0)^2)/(2*(lambda0+n)))

# Marginal posterior for mu
Uat45 <- sqrt((lambda1*alpha1)/beta1)*(0.45-mu1)
ht45 <- dt(Uat45,2*alpha1)/(sum(y1)*dmu)
Uat55 <- sqrt((lambda1*alpha1)/beta1)*(0.55-mu1)
ht55 <- dt(Uat55,2*alpha1)/(sum(y1)*dmu)
plot(mu,y1/(sum(y1)*dmu),type="l",xlab="mu",ylab="density",
  main="Posterior dstn of mu (Noninformative prior)")
lines(c(.45,.45),c(0,ht45))
lines(c(.55,.55),c(0,ht55))

# Posterior probability that Ho true
probHotrue <- pt(Uat55,2*alpha1)-pt(Uat45,2*alpha1)
probHotrue

# Compare with frequentist t-test for mu
t.test(x,mu=0.50)
```

