---
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---
```{r,include=FALSE}
library(mosaic)
```
# 9.8 Bayes Test Procedures

```{r}
# The data - TV radiation example
x <- c(0.65,0.28,0.47,0.44,0.25,1.03,0.28,0.64,0.71,0.49)
n <- length(x)
xbar <- mean(x)
snsq <- (n-1)*var(x)
```

## TV Radiation Example (one-sample t-test)

$H_0:\mu \leq 0.50$ assume safe levels vs $H_1:\mu > 0.5$

$U=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}$

Reject $H_0$ if $U>t^{-1}_{(9)}(.95)=1.833$

# P-value Interpretation (Frequentist)
```{r}
t.test(x,mu=.5,var.equal=TRUE,alternative="greater")
```

Interpretation of p-value: There is a 38% chance of getting a sample mean as larger or larger than 0.52 just by chance when the true mean radiation level is 0.50 mR/hr.

# Loss Functions
Consider two simple hypotheses: 
$$H_0:\theta=\theta_0$$
$$H_1:\theta=\theta_1$$

$d_0$: decision to **not** reject $H_0$

$d_1$: decision to reject $H_0$

The loss, $L(\theta_i,d_i)$, that occurs when $\theta_i$ is true and decision $d_j$ is chosen is given in the table: 

| | $d_0$ | $d_1$ |
|-----|-------|-------|
| $\theta_0$ | 0 | $\omega_0$ |
| $\theta_1$ | $\omega_1$ | 0 |

# Expected Loss

- Prior probability that $H_0$ is true: $\xi_0$
- Prior probability that $H_1$ is true: $\xi_1=1-\xi_0$
- Expected loss of each test procedure $\delta$: $r(\delta)$

$$r(\delta)=\xi_0 E(Loss|\theta=\theta_0)+\xi_1 E(Loss|\theta=\theta_1)$$

$$ r(\delta)=\xi_0 \omega_0 \alpha(\delta)+\xi_1\omega_1\beta(\delta)$$

- $\alpha(\delta)$: Pr(Type I error) = $Pr(d_1|\theta=\theta_0)$
- $\beta(\delta)$: Pr(Type II error) = $Pr(d_0|\theta=\theta_1)$

# Bayes Test Procedure
A procedure $\delta$ for which $r(\delta)$ is minimized is called a **Bayes test procedure**.

E.g Reject $H_0$/choose $d_1$ if 

$$\begin{aligned}
r(d_0|x)&>r(d_1|x)\\
\omega_1 Pr(H_1 \ true |x) &>Pr(H_0 \ true|x)\\
\omega_1[1-Pr(H_0 \ ture |x)]&> \omega_0 Pr(H_0 \ true|x)\\
Pr(H_0 \ true|x)&< \frac{\omega_1}{\omega_0+\omega_1}\\
\int_{\Omega_0}\xi(\theta|x)d\theta & \leq \frac{\omega_1}{\omega_0+\omega_1}
\end{aligned}$$

Choose $d_1$ if $r(d_0|x)>r(d_1|x)$, where $r(d_i|x)=\int L(\theta,d_i)\xi(\theta|x)d\theta$

Reject $H_0$ if $\frac{\omega_1}{\omega_0+\omega_1} \geq Pr(H_0 \ true |x)=\int_{\Omega}\xi(\theta|x)d\theta$

# TV Example
We need $\xi(\mu|x)$ for Bayesian Inference

- Likelihood: $f_n(x|\mu,\tau)=\prod N(\mu,\tau)$
- prior: $\xi(\mu,\tau)=\xi(\tau)\xi(\mu|\tau)=gamma(\alpha_0,\beta_0)  N(\mu_0,\lambda_0 \tau)$

$$\xi(\mu,\tau|x) \propto f_n(x|\mu,\tau) \xi(\mu,\tau)$$

## Posterior marginal of $\mu$ (Theorem 8.6.1 p496)

$U=\left(\frac{\lambda_1\alpha_1}{\beta_1}\right)^{1/2}(\mu-\mu_1)\sim t_{2\alpha_1}$

- $\mu_1 = \left(\frac{\lambda_0}{\lambda_0+n}\right)\mu_0 + \left(\frac{n}{\lambda_0+n}\right)\bar{x}_n$
- $\lambda_1=\lambda_0 +n$
- $\beta_1 = \beta_0 + \frac{1}{2}s_n^2+\frac{n\lambda_0(\bar{x}_n-\mu_0)^2}{2(\lambda_0+n)}$
- $\alpha_1 = \alpha_0 + n/2$

# Choosing Non-Informative Prior
Recall: When $X_i\sim N(\theta,\sigma^2)$ with $\sigma^2$ known $\xi(\theta)\sim N(\mu,\nu^2)$.  We let $\nu^2 \rightarrow \infty$ to get the non informative prior.  Then $\xi(\theta|x)\sim N(\bar{X}_n,\sigma^2/n)$

### How do we pick a non-informative prior for $\mu$ and $\sigma^2$ jointly?

- for mean $\mu$ consider uniform, e.g. all intervals (a, a+h) having same prior probability
- for scale parameter $\sigma = \sqrt{\frac{1}{\tau}}$ all intervals (a, ka) having same prior probability

$$\xi(\mu) =1; \xi(\tau)=\frac{1}{\tau} \rightarrow \xi(\mu,\tau)=1 \times \frac{1}{\tau}=\frac{1}{\tau}$$ 

**Jeffreys Prior**:Jeffreys prior is proportional to the square root of the determinant of the Fisher Information matrix and is invariant to reparameterization.

# Back to Bayesian Analysis of TV Radiation

$H_0:\mu \leq \mu_0$ vs $H_1:\mu > \mu_0$

Consider Pr($H_0$ true | data) and Reject $H_0$ if small. $Pr(H_0 \ true | x)=\int_{\Omega_0}\xi(\mu|x)d\mu$

1. Non-informative Prior: $\xi(\mu,\tau)=\xi(\mu|\tau)\xi(\tau)=N(\mu_0,\lambda_0\tau) gamma(\alpha_0,\beta_0)$

For what values of $\mu_0, \lambda_0, \alpha_0, \beta_0$ would $\xi(\mu,\tau)=1/\tau$ (Jeffreys Prior)?

2. Find the values of $\mu_1, \lambda_1, \alpha_1, \beta_1$ using the non informative prior.

3. Using these posterior values show the equivalence of the Bayes Test Procedure using Jeffreys prior to the frequentist result ($U\sim t_{(n-1)}$)

4. Compare the posterior probability to p-value (using R).

# And More 

5. Repeat using informative hyper parameters for the conjugate prior relationship.

6. Show equivalence with non informative prior to the two-sided t-test. ($\mu_0 = .25; \lambda_0 = 4; \alpha_0 = 2; \beta_0 = 0.5$)


