---
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---
```{r,include=FALSE}
library(mosaic)
```

# 9.1 Testing Hypotheses

Example: National Assessment of Adult Literacy (NAAL) survey scores (0-500)

- 233: can add numbers of 2 checks on a bank deposit slip
- 325: can price a meal from a menu
- 375: can trasnform from cents/oz to dollar/lb

275: can Balance a check book

We collect data from $n=840$ men ages 21-25 (assume $\sigma=59, \bar{X}_n=272$)

# Hypothesis Test Assumptions

$$X\sim N(275,59^2)$$

pvalue$=Prob(\bar{X}_n\leq 272|X\sim N(275,59^2))$

$$H_0:\mu \geq 275$$
$$H_A:\mu < 275$$

our test statistic under the $H_0$: $\bar{X}_n\sim N(275,59^2/840)$

$z=\frac{\bar{x}-\mu}{\sigma/n}\sim N(0,1)$ under $H_0$

$z_{obs}=-1.474$ and the pvalue (one sided) is 0.0708.


# Using R
```{r,fig.height=2,fig.width=2}
xpnorm(272,275,59/sqrt(840))
pnorm(272,275,59/sqrt(840))
```


# Applied to 9.1

$$\begin{aligned}
H_0 & = \theta \in \Omega_0 \\
H_1 & = \theta \in \Omega_1 \\
\end{aligned}$$

Where $\Omega_0 \cup \Omega_1 =\Omega:$ the entire sample space for $\theta$

Our case the context of 9.1
$$\begin{aligned}
\Omega_0 & = [275,500] \\
\Omega_1 & = [0,275) \\
\end{aligned}$$

Test statistic: $T=r(x)$

Critical region: Reject $H_0$ if $T\leq -1.645$

# R
```{r,fig.height=2,fig.width=2}
xqnorm(.05)
```

# Power Function

Def: The probability that a test procedure $\delta$ rejects $H_0$ for each value of $\theta \in \Omega$

$\pi(\theta|\delta)=Pr(x\in C|\theta)$ for $\theta \in \Omega$

An ideal power function:

$$\begin{aligned}
\pi(\theta|\delta) & = 0 \ \mbox{for} \ \theta \in \Omega_0 \\
\pi(\theta|\delta) & = 1 \ \mbox{for} \ \theta \in \Omega_1 \\
\end{aligned}$$

BUT probabilities of Error exist.

# Errors In Hypothese thests

| | $H_0$ True | $H_0$ False |
|-----------------|----------------|--------------------|
Accept $H_0$ | OK | Type II Error|
Reject $H_0$ | Type I Error | OK |

Prob(Type I Error)= Prob(Reject $H_0$ when $H_0$ is true) 

$$=\pi(\theta|\delta) \ \mbox{for} \ \theta \in \Omega_0 $$

Prob(Type II Error)= Prob(Do not reject $H_0$ when $H_0$ is false) 
$$=1- \pi(\theta|\delta) \ \mbox{for} \ \theta \in \Omega_1 $$

# Scenarios

| | Innocent | Gulity |
|-----------------|----------------|--------------------|
Acquit | OK | Type II Error|
Convict | Type I Error | OK |

## Which error is worse?
Convict an innocent person? Or let a criminal run free?

| | Drug does not work | Drug works |
|-----------------|----------------|--------------------|
Don't Approve | OK | Type II Error|
Approve | Type I Error | OK |

## Which error is worse?
Type I Error: Drug is marketed but doesn't really work.

Type II Error: Drug that could save lives is not available

# Hypothesis Testing Strategy
Set upper bound for Type I error.  Then among those tests, pick one that maximizes Power (minimizes Type II Error)

Require:
$$\pi(\theta|\delta) \leq \alpha_0 \ \mbox{for} \ \theta \in \Omega_0 $$

Size $\alpha(\delta)$ of a test $\delta$

$$\alpha(\delta)=\sup_{\theta \in \Omega_0} \pi(\theta|\delta)$$

# Checkbook example

- Find P(Type I Error)

- Find Size of the test if $\mu=275$

- Find power of the test if $\mu=270$

# Pr(Type I Error)

Pr(Type I Error)=Pr(Reject $H_0|H_0$ is tre)

$$\begin{aligned}
&= Pr(z\leq 1.645|\mu=275)\\
&=Pr(\frac{\bar{X}_n-275}{59/\sqrt{840}}<-1.645)\\
&=Pr(\bar{X}_n<271.65)\\
\end{aligned}$$

So $\delta:\{x:\bar(x)_n<271.65\}=\{x:z=\frac{\bar{X}_n-\mu}{\sigma/\sqrt{n}}\}$

# Size of the test

$$\begin{aligned}
\alpha(\delta)&=\sup_{\mu\geq275}\pi(\mu|\delta)\\
&=\sup_{\mu}Pr(\bar{X}_n<271.65|\mu\geq275)\\
&=Pr(\bar{X}_n<271.65|\mu=275)\\
&=Pr(Z<\frac{271.65-275}{59/\sqrt{840}})\\
&=Pr(Z<-1.645)=0.05\\
\end{aligned}$$

# Power of the test for $\mu$=270

$$\begin{aligned}
\pi(\mu|\delta) &=Pr(\bar{X}_n<271.65|\mu=270)\\
&=Pr(Z<\frac{271.65-270}{59/\sqrt{840}})\\
&=Pr(Z<0.81)=0.791\\
\end{aligned}$$

```{r}
pnorm(0.81)
```

# Plot of Type I error, Type II Error, and Power
```{r,echo=FALSE,warning=FALSE,fig.height=3,fig.width=4}
funcShaded <- function(x) {
    y <- dnorm(x, mean = 275, sd = 59/sqrt(840))
    y[x < 260 | x > 271.65] <- NA
    return(y)
}

funcShaded2 <- function(x) {
    y <- dnorm(x, mean = 270, sd = 59/sqrt(840))
    y[x < 260 | x > 271.65] <- NA
    return(y)
}
funcShaded3 <- function(x) {
    y <- dnorm(x, mean = 270, sd = 59/sqrt(840))
    y[x < 271.65 | x > 275] <- NA
    return(y)
}

myplot <- ggplot(data.frame(x = c(260,285)), aes(x = x)) +
        stat_function(fun = dnorm, args = list(275, 59/sqrt(840)),
                      aes(colour = "mean=275"), size = 1.5) +
        stat_function(fun = dnorm, args = list(270, 59/sqrt(840)),
                      aes(colour = "mean=270"), size = 1.5) +
        scale_x_continuous(name = "Probability",
                              breaks = seq(0, 1, 0.2),
                              limits=c(260,285)) +
        scale_y_continuous(name = "Frequency") +
        ggtitle("Null and Alternative ") +
        scale_colour_brewer(palette="Accent") +
        labs(colour = "Groups") + 
   stat_function(fun=funcShaded2, geom="area", fill="#84CA72", alpha=0.2) +
   stat_function(fun=funcShaded, geom="area", fill="#666666", alpha=0.2) +
   stat_function(fun=funcShaded3, geom="area", fill="#FDC086", alpha=0.2)

myplot

```


# How do the following effect Power

Power=1-Pr(Type II Error) = Probability of corretly rejeting $H_0$

- $\alpha$: as $\alpha$ increases, Power increases
- $\sigma$: as $\sigma$ decreases, Power increases
- n: as n increases, Power increases
- $\mu$: as $\mu$ decreases(futher away from $\mu_0$), Power increases

