---
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---
```{r,include=FALSE}
library(mosaic)
```

# 9.1 Testing Hypotheses: Uniform Example

$$X\sim Unif(0,\theta)$$

We want to test

$$H_0:\theta \geq 2$$
$$H_A:\theta<2$$

with $\alpha_{\theta}=0.10$ based on a random sample of size 8. We'll base our test on $\hat{\theta}_{MLE}=max(X_1,...,X_8)=X_{(8)}$.  We want to find $\pi(\theta|\delta)$.

Reject $H_0$ if $X_{(8)}\leq C$ (since we are looking for evidence that $\theta < 2$). 

We want Pr(Reject $H_0|H_0$ true)=.10 or $Pr(X_{(8)}\leq C|\theta=2)=.10$

# Finding the pdf for an order statistic

First we need the pdf for $X_{(8)}$.  Begin with cdf of $X_{(8)}$.


$$\begin{aligned}
F_{X_{(8)}}(x)&=\left[ Pr(X\leq x)\right]^8\\
f_{X_{(8)}}(x) &= 8 F(x)^{n-1}f(x)\\
& = 8 \left(\frac{x}{\theta}\right)^7\frac{1}{\theta}\\
& = \frac{8\theta^7}{\theta^8}\\
\end{aligned}$$

If $\theta=2$ then $f_{X_{(8)}}(x)=\frac{1}{32}x^7$ for $0\leq x \leq 2$.

# Finding the Critical Region
So 
$$\begin{aligned}Pr(X_{(8)}\leq C|H_0 \mbox(true)=.10 & \rightarrow \int_0^c\frac{1}{32}x^7dx=.10\\
&\rightarrow \frac{1}{256} x^8 \Big|_0^c\\
& \rightarrow c=(256)^{1/8}=1.50
\end{aligned}$$

So the Critical Region (Rejection Region) is $\{x:x_{(8)}\leq 1.50\}$.

# What is the power of the test when $\theta=1.7$?

$$Pr(\mbox{Reject} H_0|\theta=1.7) = \int_0^{1.5} \frac{8}{1.7^8}x^7dx=$$

```{r,echo=FALSE}
theta=1.7
myf<-function(x){x^7}
val<-integrate(myf, 0, 1.5)$value
8/theta^8*val
```

```{r,echo=FALSE}
theta<-c(1.5,1.6,1.7,1.8,1.9,2.0,2.1)
val<-integrate(myf, 0, 1.5)$value
y<-8/theta^8*val
y
```

|$\theta$|Power|
|---|--------|
|1.5| 1.00 |
|1.6| 0.597|
|1.7| 0.367|
|1.8| 0.233|
|1.9| 0.151|
|2.0| 0.100|
|2.1| 0.068|

# Plot of Power vs $\theta$

```{r,echo=FALSE,fig}

plot(y~theta,type="l")
abline(h=.1,lty=3)
```

# Normal Example

$X\sim N(\mu,\sigma^2)$ with $\sigma^2$ known.

$$H_0:\mu=\mu_0; H_A:\mu \neq \mu_0$$

$\pi(\mu|\delta)=Pr(\mbox{Reject} \ H_0|\mu)$

Let $\delta$ be the test that rejects when $|Z|>1.96$
\small
$$\begin{aligned}
&= Pr\left(\frac{\bar{X}_n-\mu_0}{\sigma/\sqrt{n}}>1.96|\mu\right)+Pr\left(\frac{\bar{X}_n-\mu_0}{\sigma/\sqrt{n}}<-1.96|\mu\right)\\
&= Pr\left(\bar{X}>\mu_0+1.96\times \sigma/\sqrt{n}|\mu\right)+Pr\left(\bar{X}<\mu_0-1.96\times \sigma/\sqrt{n}|\mu\right)\\
&= Pr\left(Z>\frac{\mu_0+1.96\times \sigma/\sqrt{n}-\mu}{\sigma/\sqrt{n}}\right)+Pr\left(Z<\frac{\mu_0-1.96\times \sigma/\sqrt{n}-\mu}{\sigma/\sqrt{n}}\right)\\
&= Pr\left(Z>\frac{\mu_0-\mu}{\sigma/\sqrt{n}}+1.96\right)+Pr\left(Z<\frac{\mu_0-\mu}{\sigma/\sqrt{n}}-1.96\right)
\end{aligned}$$

\normalsize

# Bernoulli Example (Example 9.1.9)

$$X_1,...,X_{10}\sim \mbox{Bernoulli}(p)$$

$$H_0:p \leq .3$$
$$H_A:p >.3$$
Test statistic: $Y=\sum X_i \rightarrow$ Reject $H_0$ if $Y>C$ where 
$$Y\sim Binomial(n,p); \ n=10$$

# Size of Binomial Test

$$\alpha(\delta)=\sup_{p \in \Omega_0}\pi(p|\delta)=\sup_{p in \Omega_0} Pr(Y\geq c|p) = Pr(Y\geq c|p=0.3)$$
In general $Pr(Y=y|p=.3)= {10 \choose y}.3^{y}.7^{10-y}$
$$\begin{aligned}
Pr(Y=10|p=.3)&= {10 \choose 10}.3^{10}.7^{0} = 0.0000059\\
Pr(Y=9|p=.3)&= {10 \choose 9}.3^{9}.7^{1}= 0.00014\\
\end{aligned}$$

# Using R
```{r}
y=seq(5,10)
d=dbinom(y,10,.3)
p=1-pbinom(y-1,10,.3)
cbind(y,d,p)
```

In order to keep the size of the test to at most .1 (or 0.05 for that matter) we must choose c>5.

# P-value

P-value: The smallest level $\alpha_0$ such that we would reject the null hypothesis at a level $\alpha_0$ with the observed data.

If $\delta=$ Reject $H_0$ when $T\geq c$
($\delta_t:$ test that rejects $H_0$ if $T\geq t$.

$$\mbox{pvalue}=\sup_{\theta \in \Omega_0}\pi(\theta|\delta_t)=\sup_{\theta \in \Omega_0}Pr(T\geq t|\theta)$$

## Bernoulli Example

If we observe $Y=8$ then 
$$\begin{aligned}
\mbox{pvalue} &=\sup_{p \leq .3}Pr(Y\geq 8|p)\\
&= Pr(Y\geq 8|p=.3)\\
&= 0.0016\\
\end{aligned}$$

# Equivalence of Hypthesis test and Confidence Intervals

A coefficient $\gamma$ confidence set can be thought of as a set of $H_0$s that would be accepted at significance level $1-\gamma$.

## Example 9.1.15

$X_1...,X_n \sim N(\mu,\sigma^2)$ with $\mu,\sigma^2$ unknown.

$$H_0:\mu \leq \mu_0; \{\Omega_0: \mu \leq \mu_0 \ \mbox{and}\  \sigma^2>0\}$$
$$H_A:\mu > \mu_0; \{\Omega_1: \mu > \mu_0 \ \mbox{and}\  \sigma^2>0\}$$
Coefficient $\gamma$ Confidence Interval:

$$\left(\bar{X}_n \pm c \sigma'/\sqrt{n} \right)$$

where $c$ is the $\frac{1+\gamma}{2}$ quantile of $t_{n-1}$ distribution.

# Normal CIs
We can use the CI to find a level $\alpha_0=1-\gamma$ test of $H_0:\mu=\mu_0; H_A:\mu \neq \mu_0$

Reject $H_0$ if $\mu_0 \notin \left(\bar{X}_n \pm c \sigma'/\sqrt{n} \right)$

Which is equivalent to $\left(|\bar{X}_n-\mu_0| \geq \pm c \sigma'/\sqrt{n} \right)$

Or $$ \Big| \frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}\Big| \geq c$$

Which looks like a t-test!
