---
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---
```{r,include=FALSE}
library(mosaic)
```

# 9.1 (continued) P-value

P-value: The smallest level $\alpha_0$ such that we would reject the null hypothesis at a level $\alpha_0$ with the observed data.

If $\delta=$ Reject $H_0$ when $T\geq c$
($\delta_t:$ test that rejects $H_0$ if $T\geq t$.

$$\mbox{pvalue}=\sup_{\theta \in \Omega_0}\pi(\theta|\delta_t)=\sup_{\theta \in \Omega_0}Pr(T\geq t|\theta)$$

## Bernoulli Example

If we observe $Y=8$ then 
$$\begin{aligned}
\mbox{pvalue} &=\sup_{p \leq .3}Pr(Y\geq 8|p)\\
&= Pr(Y\geq 8|p=.3)\\
&= 0.0016\\
\end{aligned}$$

# Equivalence of Hypthesis test and Confidence Intervals

A coefficient $\gamma$ confidence set can be thought of as a set of $H_0$s that would be accepted at significance level $1-\gamma$.

## Example 9.1.15


Coefficient $\gamma$ Confidence Interval:

$$\left(\bar{X}_n \pm c \sigma'/\sqrt{n} \right)$$

where $c$ is the $\frac{1+\gamma}{2}$ quantile of $t_{n-1}$ distribution.

# Normal CIs
We can use the CI to find a level $\alpha_0=1-\gamma$ test of $H_0:\mu=\mu_0; H_A:\mu \neq \mu_0$

Reject $H_0$ if $\mu_0 \notin \left(\bar{X}_n \pm c \sigma'/\sqrt{n} \right)$

Which is equivalent to $\left(|\bar{X}_n-\mu_0| \geq \pm c \sigma'/\sqrt{n} \right)$

Or $$ \Big| \frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}\Big| \geq c$$

Which looks like a t-test!

# Chapter 9.5 The t-test

Again, consider $X_1...,X_n \sim N(\mu,\sigma^2)$ with $\mu,\sigma^2$ unknown.

$$H_0:\mu \leq \mu_0; \{\Omega_0: \mu \leq \mu_0 \ \mbox{and}\  \sigma^2>0\}$$
$$H_A:\mu > \mu_0; \{\Omega_1: \mu > \mu_0 \ \mbox{and}\  \sigma^2>0\}$$

Reject $H_0$ if $U\geq c$ where $U=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}$ and $c=T'_{(n-1)(1-\alpha_0)}$ since $U\sim t_{(n-1)}$ when $H_0:\mu=\mu_0$ is true.

# Example TV Radition

It's 1980 and we are interested in measuring the radiation level of TV displays in a department store. The safety limit established by the [FDA](https://www.fda.gov/radiation-emitting-products/resources-you-radiation-emitting-products/television-radiation#6) is 0.5 milliroentgen per hour (mR/hr).  We collect data.

$$\begin{aligned}
n&=10\\
\bar{x}_n&=0.52\\
\sigma'&=.241
\end{aligned}$$

## Confidence Interval for $\mu$

$$\bar{x}_n \pm t^{-1}_{(n-1)}(.975)\sigma'/\sqrt{n}$$
$$.52 \pm 2.2622 \times .241/\sqrt{10}$$
$$(0.35,0.69)$$

# Hypothesis Test for TV radiation

$H_0:\mu \leq 0.50$ assume safe levels vs $H_1:\mu > 0.5$

$U=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}=\frac{.52-.50}{.241/\sqrt{10}}=.26$

Reject $H_0$ if $U>t^{-1}_{(9)}(.95)=1.833$

$$\begin{aligned}
\mbox{p-value} &=\sup_{\mu \leq 0.50}P(U\geq.26|\mu)\\
&=P(U \geq .26|\mu=0.50)=.40\\
\end{aligned}$$

Do not Reject $H_0$.  No significant evidence that unsafe levels exist.

```{r}
qt(.975,9)
1-pt(.26,9)
```

# Assumptions 

* Population is Normally Distributed
* Random sample (iid)
* Normal distribution (or n is "large enough")

## Question
How much power did this test have to reject $H_0$ when, say, $\mu=0.60$?

In general, was this test powerful enough to detect meaningful safety violations?

test statistic: $U=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}} \sim t_{(n-1)}$ if $H_0:\mu=\mu_0 \ \mbox{is true}$

But $U=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}} \not \sim t_{(n-1)}$ if $H_0$ is false. 

# Non-Central t-distribution

Definition (9.5.1): If $Y$ and $W$ are independent random variables, with $W\sim N(\psi,1)$ and $Y\sim \chi^2_{(m)}$, then the distribution of $$X=\frac{W}{\sqrt{Y/m}}$$ is called a non-central t distribution with $m$ degrees of freedom and non-centrality parameter $\psi$

## Consider U when $H_0$ is false.

$$\begin{aligned}
U&=\frac{\bar{X}_n-\mu_0}{\sigma'/\sqrt{n}}= \frac{\frac{\bar{X}_n-\mu_0}{\sigma/\sqrt{n}}}{\sqrt{\frac{(n-1)\sigma'^2}{\sigma^2}/(n-1)}}\\
&=\frac{\frac{\bar{X}_n-\mu}{\sigma/\sqrt{n}}+\frac{\mu-\mu_0}{\sigma/\sqrt{n}}}{\sqrt{\frac{\chi^2_{(n-1)}}{(n-1)}}}=\frac{N(\psi,1)}{{\sqrt{\chi^2_{(n-1)}/(n-1)}}}\sim t_{(n-1),\psi}
\end{aligned}$$

# Calculating Power

Then the power at $\mu$ is 
$$1-T_{(n-1)(c | \psi)}=\int_{1.833}^{\infty} f_T(t)dt$$ 

* $P(X\leq t)=T_{n}(t|\psi)$ 

* $c=$ critical value

* Reject $H_0$ if $U\geq c=1.833$

* $\psi=\frac{\mu-\mu_0}{\sigma/\sqrt{n}}=\frac{.60-.50}{\sigma/\sqrt{10}}$

## But we need $\sigma$...

# Power Using Non-Central t

1. Find power at $\mu=.6$ assuming $\sigma=s$ (sample standard deviation)
2. Find power as a function of effect size in standard deviations (express $\mu$ in terms of $\sigma$ and $\mu_0$)
3. Assuming $\mu=\mu_0+1\times \sigma$, plot the following
     a. Power vs $\psi$
     b. Power vs $n$
     

# Find power at $\mu=.6$ assuming $\sigma=s$ (sample standard deviation)

```{r}
psi=(.6-.5)/(.241/sqrt(10))
1-pt(1.833,df=9,ncp=psi)
```

# Find power as a function of effect size in standard deviations (express $\mu$ in terms of $\sigma$ and $\mu_0$)

```{r}
psi=1*sqrt(10)
1-pt(1.833,df=9,ncp=psi)
```

# Bad plot trying to look at different number of SDs away from null
```{r,echo=FALSE,warning=FALSE}
k=c(seq(-2,2,by=.5))
psi=k*sqrt(10)
power=1-pt(1.833,df=9,ncp=psi)
plot(power~k)

```

# Assuming $\mu=\mu_0+1\times \sigma$: Power vs $\psi$

```{r}
n=c(seq(3,50,by=1))
psi=.5*sqrt(n)
power=1-pt(1.833,df=(n-1),ncp=psi)
plot(power~psi)
```

# Power vs n
```{r}
plot(power~n)
```


