---
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---
```{r,include=FALSE}
library(mosaic)
```

```{r,echo=FALSE}
clouds<-read.csv("~/Stats 322 S20/Class/Data/CloudSeeding.csv")
clouds$lograin=log(clouds$rainfall)
```


# 9.6 Two Sample t-test
We will look at an example comparing means of two Normal Distributions.  We will explore some of the properties, investigate power, and consider whether this is a LRT.

## Example: Cloud Seeding (Example 8.3.1 & 9.6.1)

Comparing log rainfall over the two groups

\tiny
```{r}
favstats(lograin~group,data=clouds)
```

\normalsize

# Is there evidence of greater rainfall in the seeded group compared to the unseeded group?
Is there a significant difference in the average amount of rainfall?

* hypothesis test/p-value
* Confidence Intervals
* Interpretations/Conclusions (Frequentist perspective)
* Assumptions

# Set-up

$X=(X_1,...,X_m)\sim N(\mu_1,\sigma^2); Y=(Y_1,...,Y_n)\sim N(\mu_2,\sigma^2)$

$H_0:\mu_1 \leq \mu_2; H_1:\mu_1>\mu_2$

Note: $\sigma^2$ unknow and also the same in each group (equal variance assumption)

$$\mbox{test statistic} =\frac{\mbox{estimate - hypothesized value}}{\mbox{standard error of the estimate}}$$

$$U=\frac{(\bar{X}_m-\bar{Y}_n)-0}{SE_{(\bar{X}_m-\bar{Y}_n)}}$$

# Standard Error

$$\begin{aligned}
Var(\bar{X}_m-\bar{Y}_n) & = Var(\bar{X}_m) + Var(\bar{Y}_n)\\
& = \frac{\sigma_X^2}{m}+\frac{\sigma_Y^2}{n}\\
sd(\bar{X}_m-\bar{Y}_n) &= \sqrt{ \frac{\sigma_X^2}{m}+\frac{\sigma_Y^2}{n}}
\end{aligned}$$

If $\sigma_X^2=\sigma_Y^2=\sigma^2$ as we've assumed, then our best estimate of $\sigma^2$ is
$$s_p^2=\frac{(m-1)s_x^2+(m-1)s_y^2}{m+n-2}=\frac{\sum (X_i-\bar{X}_m)^2+\sum(Y_i-\bar{Y}_n)^2}{(m-1)+(n-1)}$$

So $SE(\bar{X}_m-\bar{Y}_n)=\sqrt{\frac{s_p^2}{m}+\frac{s_p^2}{n}}=s_p \sqrt{\frac{1}{m}+\frac{1}{n}}$

# Two Sample Test Statistic

$$U=\frac{(\bar{X}_m-\bar{Y}_n)}{s_p \sqrt{\frac{1}{m}+\frac{1}{n}}}$$

## Claim: U follows a $t_{(m+n-2)}$ distribution when $H_0$ is true ($\mu_1=\mu_2$)

$$U=\frac{\frac{(\bar{X}_m-\bar{Y}_n)}{\sigma\sqrt{\frac{1}{m}+\frac{1}{n}}}}{\sqrt{s_p^2/\sigma^2}}$$

Numerator is $Z\sim N(0,1)$ as desired.  Denominator can be written as: 

$$\sqrt{\frac{1}{\sigma^2} \frac{(m-1)s_x^2+(n-1)s_y^2}{m+n-2}}=\sqrt{\frac{\frac{(m-1)s_x^2}{\sigma^2}+\frac{(n-1)s_y^2}{\sigma^2}}{(m+n-2)}}$$

# Distribution of U
$\frac{(m-1)s_x^2}{\sigma^2}\sim \chi^2_{(m-1)}; \frac{(n-1)s_y^2}{\sigma^2}\sim \chi^2_{(n-1)}$

So $$U=\frac{Z}{\sqrt{\chi^2_{(m+n-2)}/(m+n-2)}} \sim t_{(m+n-2)}$$

## Further

* Reject $H_0$ if $U > T^{-1}_{(m+n-2)}(1-\alpha_0)$.

* pvalue $= Pr(U > T_{(m+n-2)}|\mu_1=\mu_2)$

* ($1-\alpha_0$)CI for $\mu_1-\mu_2$
$$\bar{X}_m-\bar{Y}_n \pm T^{-1}_{(m+n-2)}(1-\alpha_0/2) s_p\sqrt{\frac{1}{m}+\frac{1}{n}}$$

# Using R

\scriptsize
```{r}
m <- 26
n <- 26
xbar <- mean(clouds$lograin[clouds$group=="Seeded"])
sxsq <- (m-1)*var(clouds$lograin[clouds$group=="Seeded"])
ybar <- mean(clouds$lograin[clouds$group=="Unseeded"])
sysq <- (n-1)*var(clouds$lograin[clouds$group=="Unseeded"])
sp2<- (sxsq+sysq)/(m+n-2)
se <- sqrt(sp2)*sqrt(1/n+1/m)
U=(xbar-ybar)/se;U
t.test(lograin~group,var.equal=TRUE,data=clouds)
```

# What is the power function like?

\scriptsize
```{r}
qt(0.95,df=50)
```

Reject $H_0$ if U>1.67.  (We observed U=2.54 so we reject $H_0$.)

$$\begin{aligned}
\pi(\mu_1,\mu_2,\sigma^2|\delta)&=Pr(\mbox{Reject} H_0|\mu_1,\mu_2)\\
&=Pr(U>c|\mu1,\mu2)\\
& =Pr\left(\frac{(\bar{X}_m-\bar{Y}_n)-0}{s_p\sqrt{1/m+1/n}}>c|\mu_1,\mu_2\right)\\
&= T_{(m_+n-2)(c|\psi)}
\end{aligned}$$

Where $\psi=\frac{\mu_1-\mu_2}{\sqrt{\frac{\sigma^2}{m}+\frac{\sigma^2}{n}}}$

Consider power when $\mu1-\mu2=\Delta$; or a function of $\sigma: \mu_1=\mu_2-\sigma$

\scriptsize
```{r}
psi=sqrt(1/m+1/n)
1-pt(1.67,df=50,ncp=psi)
```


# Two-sample t-test with equal variance

- The two sample t-test is fairly robust to violations of normality and equal variances.

- But it's not so good when n and m are small and there is a hight degree of skewness

- The Two-sample t-test is a likelihood ratio test when the variances are equal (p592-3).  

## What happens if we relax the assumption of equal variance?

$X=(X_1,...,X_m)\sim N(\mu_1,\sigma_1^2); Y=(Y_1,...,Y_n)\sim N(\mu_2,\sigma_2^2)$

* If we the variance of one group is a known ratio of the variance of the other group ($\sigma_2^2=k \sigma_1^2$) then U has the same t-distribution.

* If the variances are unequal the Likelihood Ratio Test Statistic has no known distribution.  This is known as the *Behrens-Fisher problem*.  

# What do we do?

When variances are not equal (and not a known ratio of each other) we can

1. Always use the Welch test which uses approximate distributions.
2. Conduct a hypothesis test 

$$H_0:\sigma_1^2=\sigma_2^2; H_1:\sigma_1^2 \neq \sigma_2^2$$

But for this test we need an F-distribution.  

# 9.7 F Distributions

If $Y \sim \chi^2_m$ and $W \sim \chi^2_n$ where $Y$ and $W$ are independent then $X$ follows an F distribution with $m$ and $n$ degrees of freedom. (Def 9.7.1)

$$ X=\frac{Y/m}{W/n}\sim F(m,n)$$

# Details
$X=(X_1,...,X_m)\sim N(\mu_1,\sigma_1^2); Y=(Y_1,...,Y_n)\sim N(\mu_2,\sigma_2^2)$

$$\frac{(m-1)s_x^2}{\sigma_1^2}=\frac{\sum(X_i-\bar{X}_m)^2}{\sigma_1^2}\sim \chi^2_{(m-1)}$$

Similarly, $\frac{(n-1)s_y^2}{\sigma_2^2} \sim \chi^2_{(n-1)}$

So $$ V=\frac{\frac{\sum(X_i-\bar{X}_m)^2}{\sigma_1^2}/(m-1)}{\frac{\sum(Y_i-\bar{Y}_n)^2}{\sigma_2^2}/(n-1)} \sim F_{(m-1),(n-1)}$$

# Cloud Seeding Example

Under the Null:$\sigma_1^2=\sigma_2^2$ 

So $$ V=\frac{\frac{\sum(X_i-\bar{X}_m)^2}{\sigma_1^2}/(m-1)}{\frac{\sum(Y_i-\bar{Y}_n)^2}{\sigma_2^2}/(n-1)} = \frac{s_x^2}{s_y^2}\sim F_{(m-1),(n-1)}$$

Reject $H_0$ if $V<c_1=F_{(m-1),(n-1)}(\alpha_0/2)$ or $V>c_2=F_{(m-1),(n-1)}(1-\alpha_0/2)$

# The F distribution in R

\tiny
```{r,fig.width=3,fig.height=2}
varx <- var(clouds$lograin[clouds$group=="Seeded"])
vary <- var(clouds$lograin[clouds$group=="Unseeded"])
varx/vary
xqf(c(.025,.975),df1=n-1,df2=m-1)
```

Pvalue: 
```{r}
pf(.949,df1=25,df2=25)
```

No significant differnece of unequal variances.

# Properties of the F-test

- Confidence Intervals for comparing variances (ratios)

- If $X\sim F_{m-1,n-1} \frac{1}{X}\sim F_{n-1,m-1}$

- If $U\sim t_n then U^2\sim F_{1,n}$

- Power function
